Quantifying Directionality and Innovation Output in Sweden's Transition to a Forest-Based Bioeconomy

### 2025-04-25 paper.qmd
update for talk at circle

---
title: Quantifying Directionality and Innovation Output in Sweden's Transition to a Forest-Based Bioeconomy
subtitle: Draft
author:
  name: Jonas Kreutzer
  affiliation: 'Department of Economic History, Lund University'
date: last-modified
abstract: |
  The transition to a bioeconomy is high on the agenda for research, industry and policy. Like other sustainability transition, the bioeconomy transition is characterized by contestation over desired outcomes and transition pathways. While shared visions are seen as pivotal in guiding transitions, the bioeconomy discourse is marked by the contention of three competing visions. The Swedish forest-based bioeconomy is well established as a case study in this arena. Yet, knowledge about innovation produced across the entire system is scarce. The use of detailed innovation data from a Swedish literature-based innovation output database, SWINNO, allows to capture the actual innovation output of the Swedish forest-based bioeconomy innovation system. Qualitative coding of the source material links innovation from 1970 to 2021 with visions of a bioresource, biotechnology and / or bioecology bioeconomy. Findings reveal a decrease in the innovation output of the forest-based bioeconomy relative to Sweden's overall commercialized innovation. Notably, innovations associated with a bioresource vision surged during the energy crises of the 1970s, pre-dating their formalization in strategy documents, suggesting a significant role for path-dependency in shaping shared visions. Concurrently, the presence of innovations aligning with multiple visions indicates an open opportunity space for future development across various directions. This research underscores the dynamic interplay between historical contexts, shared visions, and innovation trajectories within the bioeconomy transition, and questions the extent to which visions can provide novel directionality to transitions.

csl: 'https://zotero.org/styles/apa?source=1'
bibliography:
  - swinno-shades-green.bib
link-citations: true
link-bibliography: true
colorlinks: true
number-sections: true
number-depth: 3
format:
  pdf:
    documentclass: scrartcl
    papersize: a4
    linestretch: 1.25
    mainfont: "Adobe Garamond Pro"
    sansfont: "Adobe Garamond Pro Semibold"
    indent: true
execute:
  warning: false
  error: false
  echo: false
jupyter: python3
---

```{python}
# | label: Imports

import pandas as pd
import numpy as np
import scipy.stats as stats
from pyhere import here
from IPython.display import display, Markdown
import matplotlib.pyplot as plt
import matplotlib.ticker as mtick
import seaborn as sns
from tabulate import tabulate

from src.figures import *
from src.utils import get_project_root, format_table
from src.data_handling import database

ROOT = get_project_root()
raw_data = here("data/raw")
modified_data = here("data/modified")
conn = database.connect_swinno_db()

# set all text on plots to be monospace with rcParams
plt.rcParams["font.family"] = "Space Grotesk"

sns.set_theme(style="white")
```

```{python}
#| label: Load Data
raw_data = here("data/raw")
uncleaned_bioeconomy_query = pd.read_csv(raw_data / "uncleaned_bioeconomy_query.csv")
bioeconomy = pd.read_csv(raw_data / "id_year_bioeconomy.csv")
bioeconomy = bioeconomy.loc[bioeconomy["bioeconomy"] == 1]

sni_codes = pd.read_sql_query(
    """
select
*
from 
sni_codes
""",
    conn,
)

sni_codes = sni_codes.rename(columns={"code": "use_sector"})

swinno = pd.read_sql_query(
    """
select sinno_id, year_of_commercialization as year, innovation_name_in_swedish as name
from innovation;
""",
    conn,
)

swinno["bioeconomy"] = (
    swinno["sinno_id"].astype(str).isin(bioeconomy["sinno_id"].astype(str).unique())
)

# full code labels for merging
codes_df = pd.read_sql("SELECT * FROM classification_codes", conn)

codes_df["Category"] = codes_df["Category"].str.strip()
uncertain_eco = {"Code": ["000"], "Category": ["Unsure"]}
codes_df = pd.concat([codes_df, pd.DataFrame(uncertain_eco)], ignore_index=True)

codes_df = codes_df[~codes_df["Code"].duplicated(keep="last")]

codes_df = codes_df.set_index("Code")

eco_innovations = database.query_eco_innovations(conn)
eco_innovations["innovation_type"] = eco_innovations["innovation_type"].str.strip()

bioeconomy_visions = pd.read_csv(raw_data / "bioeconomy_visions.csv")
```

```{python}
# | label: Innovation Counts
n_swinno = swinno["sinno_id"].nunique()
year_min = swinno["year"].min()
year_max = swinno["year"].max()

n_uncleaned_bioeconomy_innovations = uncleaned_bioeconomy_query["sinno_id"].nunique()
sum_bioeconomy_innovations = swinno["bioeconomy"].sum()
```

# Introduction {#sec-introduction}

<!-- Paragraphs 1-2: Motivation. After reading these paragraphs a reader in any field of economics should believe that if you answer your research question your paper will make an important contribution. -->
At the heart of sustainability transitions lies the question how societal functions can be fulfilled within ecological boundaries [@kohler2019AgendaSustainabilityTransitions].
Societal functions are provided through socio-technical systems, which consist of the interplay between the production, distribution, and use of technology [@geels2004SectoralSystemsInnovation].
Transitions are complex, long-term processes of shifts between different systems, involving diverse actors who have to chose between multiple, open-ended pathways, uncertain of their technological, social and political success [@kohler2019AgendaSustainabilityTransitions].
Further, success in the outcome and chosen pathway of sustainability transitions is contested based on the values, norms, and goals of  different actors and social groups.
Hence, innovation cannot be seen as a scalar, but must rather be understood as a vector, as possessing and providing direction to change rather than being simply progress forward [@stirling2011PluralisingProgressIntegrative].
@andersson2021OutcomesDirectionalityMorphology term these two properties _positive_ and _normative_ directionality, respectively.
Positive directionality of innovation arises from the fact that technological change is path-dependent [@arthur1989CompetingTechnologiesIncreasing;@david1985ClioEconomicsQWERTY], while normative directionality is rooted in the assumption that without public policy, private actors lack intrinsic incentives to pursue sustainability [@kohler2019AgendaSustainabilityTransitions].

A sustainability transition which has been gaining relevance for researchers, policy makers, and producers alike is the bioeconomy.
Broadly speaking, it aims to radically shift economic systems from relying on fossil fuels for energy and material resource, to an economic system built on biobased materials [@bennich2018BioBasedEconomyDynamics], circular and renewable value chains [@damato2017GreenCircularBio] and higher environmental integrity through innovation [@bugge2016WhatBioeconomyReview].
As one of the main sources for biomass in a bioeconomy [@priefer2017PathwaysShapeBioeconomy], forests have become a key arena for the bioeconomy transition, highlighting tradeoffs between material provision, carbon storage and biodiversity [@hogbom2021TrilemmaNordicBaltic].
Underneath the umbrella term "bioeconomy", the concept and viable transition pathways are heavily contested [@allain2022BioeconomicsVsBioeconomy].
Three different visions of a bioeconomy compete in discourse to provide normative directionality of the transition: the first vision centers on the provision of biomass and renewable resources, the second on the scientific development and application of biotechnology, while the third emphasizes ecological considerations[@bugge2016WhatBioeconomyReview; @vivien2019HijackingBioeconomy].
Although the discourse is well understood in terms of content [e.g., @bugge2016WhatBioeconomyReview;@vivien2019HijackingBioeconomy] its narratives [@bauer2018NarrativesBiorefineryInnovation], and actors [@holmgren2022WhoseTransformationThis], the material effects of this transition, however, remain important research topics [@fischer2020SwedishForestsBioeconomy].
How many innovations are commercialized in this new economy, especially considering the central role of innovation in bioeconomy discourse? And what does commercialized innovation suggest for the positive directionality of the bioeconomy transition?

<!-- @fischer2020SwedishForestsBioeconomy also note the consensus builidng of Swedish bioeconomy discourse marginalizes non-mainstream opinions. I can show that this is rooted, at least in part, in the path-dependence on technology output. -->

<!-- • Paragraphs 3-4: Challenges. These paragraphs explain why your research question has not already been answered, i.e., what are the central challenges a researcher must tackle to answer this question. -->
<!-- • Paragraph 5: This Paper. This paragraph states in a nutshell what the paper accomplishes and how. -->
<!-- • Paragraphs 6-7: Model. Summarize the key formal assumptions you will maintain in your analysis. -->
<!-- • Paragraphs 8-9: Data. Explain where you obtain your data and how you measure the concepts that are central to your study. -->
<!-- • Paragraphs 10-11: Methods. Explain how you take your model to the data and how you overcome the challenges you raised in paragraphs 3-4. -->

To unpack the plurality of possible socio-technical development paths, requires high data quality when measuring innovation.
Although progress towards monitoring the bioeconomy transition has been made [@jander2019MonitoringTransitionBioeconomy; @jander2020MonitoringBioeconomyTransitions], assessing directionality is limited by a lack of innovation output measures [@wydra2020MeasuringInnovationBioeconomy].
In this paper, I use SWINNO, a literature-based innovation output database capturing all significant commercialized innovation in Sweden [@sjoo2014DatabaseSwedishInnovations], to analyze the innovation output of the Swedish forest-based bioeconomy.
I develop and apply a classification scheme to identify which bioeconomy vision an innovation produced or used in forestry or the downstream value chain aligns with.

<!-- • Paragraphs 12-13: Findings. Describe the key findings. Make sure they connect clearly to the motiva-
tion in paragraphs 1-2. -->
Results show that Swedish bioeconomy innovation output has declined over the last 50 years.
Of the three different visions the bioresource vision is the most pronounced, in line with expectations from discourse analysis and the Swedish bioeconomy strategy.
However, the strategy should not be seen as visionary, as the innovation system had been producing bioeconomy innovation and even bioresource aligned innovation from the 1970s onwards.
This points to a strong interconnection between path-dependence and the creation of directionality in the policy space.

<!-- • Paragraphs 14-15: Literature. Lay out the two main ways your paper contributes to the literature. Each paragraph should center around one contribution and should explain precisely how your paper differs from the most closely related recent work. -->

<!-- By analyzing the innovations produced by the forest-bioeconomy system in Sweden, this paper adds to existing literature on directionality in sustainability transitions and bioeconomy literature in two main ways.
On one hand, using literature covering innovation output can be used to assess the directionality a sociotechnical system  -->

<!-- From an empirical perspective, this is essentially about furthering the current understanding of how sociotechnical systems grow as a result of historical path-dependencies, system-internal dynamics and external influence, towards a more elaborate understanding of how these mechanisms influence how system configurations change over time. From a conceptual perspective, it is rather about enabling such investigations, by developing analytical frameworks that capture not only the pace, but also the directionality of innovation processes. For example, there may be a need to advancethe commonly used technological innovation systems framework, since the typology of functional processes is geared towards investigating system growth in a way that hides the potential for vastly different technical, social and spatial development trajectories (Andersson, 2020).
@andersson2021OutcomesDirectionalityMorphology [127-128] -->

# Directionality and Visions of Bioeconomy {#sec-literature}

Previous literature has investigated directionality of socio-technical systems and emphasized the importance of generating shared visions as tools to provide normative directionality.
Three distinct visions of desirable transition pathways and outcomes compete for normative directionality of the bioeconomy transition.
This section introduces key concepts from this literature.

## Directionality {#sec-directionality}

Technology does not change randomly.
Instead, technological change is contingent and influences technological change at previous or later points in time.
It is path-dependent, meaning events at one point in time affect later events, which may even lead to suboptimal outcomes [@david1985ClioEconomicsQWERTY].
Path-dependency arises due to positive feedback effects, such as increasing returns [@arthur1989CompetingTechnologiesIncreasing], or through extension of possibilities adjacent to exiting technology [@kauffman2000Investigations].
Instead of more or less progress, each innovation implies a new direction for socio-technical systems [@stirling2011PluralisingProgressIntegrative].
@andersson2021OutcomesDirectionalityMorphology term this quality of innovation possessing direction _positive_ directionality.

As innovation policy took a normative turn under which innovation is needed to solve grand societal challenges [@schot2018ThreeFramesInnovation], the question of how to steer innovation into desired directions became a central issue in innovation research [@kohler2019AgendaSustainabilityTransitions].
Innovation policy now ought to promote the desired vector of societal development, instead of focusing on providing the right amount of progress [@schot2018ThreeFramesInnovation], lending innovation _normative_ directionality [@andersson2021OutcomesDirectionalityMorphology].
As innovation systems are embedded in a spatial and temporal context, not all potentially desired pathways are available or feasible. Instead, socio-technical systems are constrained by path-dependency [@kemp2022DiversifyingDeepTransitions;@bergek2015TechnologicalInnovationSystems; @stirling2011PluralisingProgressIntegrative], sometimes to the point of being locked-in to specific system states [@unruh2000UnderstandingCarbonLockin].

Dealing with the question of how to provide and exercise normative transition directionality  Weber and Rohracher [-@weber2012LegitimizingResearchTechnology] synthesize policy insights from innovation systems and multilevel perspective research into a failures framework.
Therein, directionality failure is understood as the failure to provide direction to the process of search and unclear or missing setting of collective priorities.
They caution that it should not be confused with shortsighted or flawed anticipation of market developments, as this myopia lacks a view on the direction of change.
To overcome this failure, that is, to provide directionality, they argue, external requirements need to be internalized by the innovation system in a interpretive process of negotiation between different system actors.
Reaching this collective understanding of direction, Weber and Rohracher [-@weber2012LegitimizingResearchTechnology p. 1042] write, is in practice, "... achieved by establishing shared future visions".

In the two central research streams of sustainability transitions, the innovation systems and the multi-level perspective literature, the importance of visions is highlighted.
@hekkert2007FunctionsInnovationSystems provide a framework of intervention in innovation on a systemic level, which highlights the importance of guidance of search as one of seven functions innovation system must provide.
They explain guidance of search as, "those activities within the innovation system that can positively affect the visibility and clarity of specific wants among technology users" [@hekkert2007FunctionsInnovationSystems, 423].
Changing preferences in society, decides which direction of technological development is most promising through market and government influence and, ".. an interactive and cumulative process of exchanging ideas between technology producers, technology users, and many other actors" [@hekkert2007FunctionsInnovationSystems, 423]. 
@bauer2017TechnologicalInnovationSystems adapts the innovation system functions to the specific case of biorefineries in Sweden and define the influence on the direction of search as "the incentives and/or pressures for organizations to enter the technological field. These may come in the form of visions, expectations of growth potential, regulation, policy targets, standards, articulation of demand from leading customers, crises in current business, etc." (p. 4).
Using the multi-level perspective, @yang2022ShapingDirectionalitySustainability too highlight the importance of visions in directionality over transformations, stressing the importance of shared visions between niche and regime actors.
Similarly, @bolton2019GrandVisionsPragmatic illustrate how visions and pragmatic change combined in the development and evolution of a highly integrated European electricity system.

@rotmans2001MoreEvolutionRevolution even go so far to argue that transition management is based on appealing and imaginative long-term visions and @loorbach2010TransitionManagementSustainable argues that shared visions need to be developed and disseminated between transition stakeholders.
But, as @bolton2019GrandVisionsPragmatic show, visions alone are insufficient to provide directionality.
In addition to being appealing, visions must also be specific and actors need to buy into them [@grillitsch2019InnovationPolicySystemwidea].
In summary, visions can be seen as shared imaginaries, mental models and aspirations of desired transitions which are expressed through narratives in discourse between actors [@bauer2018NarrativesBiorefineryInnovation; @holmgren2022WhoseTransformationThis].

However, one transition can be characterized by multiple, competing visions, as is illustrated by the transition to a bioeconomy.
The next section sharpens the conceptual understanding of what is meant with a bioeconomy and introduces the three visions.

## Clarification of Bioeconomy Concept

Although the bioeconomy has gained importance as a research, policy and industry focus, it remains unclear what exactly it entails.
At a high level, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources [@bugge2016WhatBioeconomyReview].
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders [@dieken2021MultitudesBioeconomiesSystematic]. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficent wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other eco-system services such as forest biodiversity.
Resolving these tradeoffs is challenging, as goal conflicts between different development pathways make simulatneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

As one of the leading forestry nations in the world, Sweden sees substantial economic and political potential in advancing the bioeconomy .
In a report commissioned by the Swedish Government, @formas2012SwedishResearchInnovation lays out the plan for a Swedish bioeconomy based on two central pillars.
It is to be an economy based on,

@@

While the official Swedish bioeconomy strategy is still under development, the content of the quotes above echos across policy documents and the national forest program [@fischer2020SwedishForestsBioeconomy].
These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread, dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic] and form the main industrial visions [@befort2020GoingDefinitionsUnderstand]. 
A third vision, centered around ecology and an economy compatible with the biosphere acts as a counter point.

@@

In summary, despite -- or perhaps because of -- bioeconomy lacking a strong definition, the term is used to refer to different concepts, with different aims and objectives, foci, normative assumptions and hence implications for society and environment [@bugge2016WhatBioeconomyReview]. The summarized visions express themselves as narratives, actor coalitions and discourse artefacts in attempts to provide normative directionality to the transition.
This normative directionality is expected to translate into positive directionality of technological change through innovation.

# Data and Methods {#sec-data_methods}

>>>>

---
title: Quantifying Directionality and Innovation Output in Sweden's Transition to a Forest-Based Bioeconomy
subtitle: Draft
author:
  name: Philipp Jonas Kreutzer
  email: philipp_jonas.kreutzer@ekh.lu.se
  orcid: 0000-0003-4234-9043
acknowledgements: |
  I am thankful for helpful comments from Josef Taalbi, Astrid Kander, Frederic Bauer as well as participants of the 9th NEST Conference and the Lund Economic History Conference Seminar Series.
affiliations: 
  - id: LU
    department: Department of Economic History
    name: Lund University
    city: Lund
    country: Sweden
funding: This work has been funded by Vinnova (2020-01963).
date: last-modified
abstract: |
  The transition to a bioeconomy is high on the agenda for research, industry and policy. Like other sustainability transitions, the bioeconomy transition is characterized by contestation over desired outcomes and transition pathways. While shared visions are seen as pivotal in guiding transitions, the bioeconomy discourse is marked by the contention of three competing visions. The Swedish forest-based bioeconomy is well established as a case study in this arena. Yet, knowledge about innovation produced across the entire system is scarce. The use of detailed innovation data from a Swedish literature-based innovation output database, SWINNO, allows to capture the actual innovation output of the Swedish forest-based bioeconomy innovation system. Qualitative coding of the source material links innovation from 1970 to 2021 with visions of a bioresource, biotechnology and/or bioecology bioeconomy. Findings reveal a decrease in the innovation output of the forest-based bioeconomy relative to Sweden's overall commercialized innovation. Notably, innovations associated with a bioresource vision surged during the energy crises of the 1970s, pre-dating their formalization in strategy documents, suggesting a significant role for path-dependency in shaping shared visions. Concurrently, the presence of innovations aligning with multiple visions indicates an open opportunity space for future development across various directions. This research underscores the dynamic interplay between historical contexts, shared visions, and innovation trajectories within the bioeconomy transition, and questions the extent to which visions can provide novel directionality to transitions.
keywords: 
  - Directionality
  - Bioeconomy
  - Forest
  - Transformative Innovation Policy
bibliography:
  - swinno-bioeconomy-directionality.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Environmental Innovation and Societal Transitions
      formatting: review
      model: 3p
      cite-style: authoryear
      highlights:
        - Highlight 1
        - Highlight 2
        - Highlight 3
execute:
  warning: false
  error: false
  echo: false
jupyter: python3
---

{{< include _code_setup.qmd >}}

<!-- TODO there are 6 unsure bioeconomy innovations remaining -->

<!-- Paragraphs 1-2: Motivation. After reading these paragraphs a reader in any field of economics should believe that if you answer your research question your paper will make an important contribution. -->
At the heart of sustainability transitions is the idea that some configurations of sociotechnical systems are preferable to others.
At least since the normative turn of innovation policy, this has gone hand in hand with the idea that policymakers ought to steer sociotechnical transitions into preferred directions.
A key instrument literature proposes for providing this normative directionality are shared visions among actors.
A central question is how these visions should emerge, should they be imposed top-down, or develop from the bottom-up?

A sustainability transition which has been high on the agenda of policymakers and researchers alike is the transition to a bioeconomy.
The fundamental idea of this transition is to move from an economic system which depends on fossil resources for fuels and raw material input, to one that builds on renewable biomass instead.
However, beyond this very broad understanding of the bioeconomy, the what, how, who, when and why of the bioeconomy transition is heavily contested.
Different actor coalitions have proposed, diffused and countered three competing visions of bioeconomies.
Expectations on future bioeconomy, and hence research on bioeconomy transition, are especially pronounced in countries endowed in biomass sources.
As a country whose land area is covered to 70% with forests and a technology leader, Sweden places high expectations on the transition to a forest-based bioeconomy and hence is well explored as a case study on the bioeconomy transition.
The central aims of this paper are to understand the extent to which commercialized innovation aligns with the contested bioeconomy visions.
Based on this case study, the paper also aims to provide insights on how innovation output and directionality through visions interact.

<!-- Paragraphs 3-4: Challenges. These paragraphs explain why your research question has not already been answered, i.e., what are the central challenges a researcher must tackle to answer this question. -->

Previous research on directionality in bioeconomy transitions has relied heavily on discourse analysis as it is challenging to identify transition outcomes of an ongoing, or expected transition.
Assessing the extent to which visions are formed bottom-up or top-down is hindered by the requirement for clearly delineated visions for each transition.

<!-- Paragraph 5: This Paper. This paragraph states in a nutshell what the paper accomplishes and how. -->

This paper overcomes these issues by turning to innovation output data, obtained from a literature-based innovation output method database of significant Swedish innovation, and classifying innovation commercialized between 1970 and 2021 into corresponding visions.
Thus, it becomes possible to observe the extent to which innovation producers have pursued different directions.
From the results it becomes evident that the vision dominating discourse about future bioeconomy configurations by key actors is also the vision which most innovations were aligned with.
However, considering that the discourse around the bioeconomy transition has only in recent years gained substantial attention resulting in the definition of what each vision entails, it seems more likely that the dominating vision formed around innovations already produced by key actors, rather than this vision providing top-down directionality.
Qualitative analysis of the innovations further supports the idea that economic and political incentives had more substantial impacts on innovation output than collectively shared understandings about ideal innovation outcomes.
This raises questions about the effectiveness of vision as instruments for translating normative directionality into innovation outcomes and echos recent work on the intertwined nature of directionality policy [@degraaff2025PoliticsDirectionalityInnovation].
By analyzing the vehciles of technological change, commercialized innovation, it also fills a gap in our understanding of directionality in sociotechnical systems [@yang2022ShapingDirectionalitySustainability].

<!-- Paragraphs 6-7: Model. Summarize the key formal assumptions you will maintain in your analysis. -->

<!-- Paragraphs 8-9: Data. Explain where you obtain your data and how you measure the concepts that are central to your study. -->

<!-- Paragraphs 10-11: Methods. Explain how you take your model to the data and how you overcome the challenges you raised in paragraphs 3-4. -->

<!-- Paragraphs 12-13: Findings. Describe the key findings. Make sure they connect clearly to the motivation in paragraphs 1-2. -->

<!-- Paragraphs 14-15: Literature. Lay out the two main ways your paper contributes to the literature. Each paragraph should center around one contribution and should explain precisely how your paper differs from the most closely related recent work. -->

# Previous Literature {#sec-literature}

Previous literature has investigated directionality of sociotechnical systems and emphasized the importance of generating shared visions as tools to provide normative directionality.
This section introduces key concepts and open questions from this literature.

## Key concepts: 
### Directionality

Increasingly scholars have paid attention to opening up innovation from notions of one directional progress or advancement of society to acknowledging that

#### Positive Directionality
#### Normative Directionality

## Open Questions
How directionality is:
        a. set, 
        b. translated,
        c. if it is (or ought to be) a top-down or bottom-up process

Directionality in the (forest) bioeconomy.

1. Concept of the bioeconomy
2. Contested Visions
    As is often the case in sustainability transitions [@wanzenbock2020FrameworkMissionorientedInnovation], the directionality of the bioeconomy is highly contested [@vivien2019HijackingBioeconomy; @bugge2016WhatBioeconomyReview].
Three distinct visions of desirable transition pathways and outcomes compete for normative directionality of the bioeconomy transition.

## Directionality {#sec-directionality}
<!-- “Through this co-evolution, transitions unfold as complex, open-ended processes. Importantly, this implies that transitions can take more and less desirable courses (Rotmans, 2005; Grin et al., 2010). In recent years, transitions scholarship has become more sensitive to this directionality of transitions, i.e. the diversity of possible transition pathways (Stirling, 2011, 2019; Røpke, 2012; Pel et al., 2020b; Andersson et al., 2021).” ([Pel et al., 2023, p. 7](zotero://select/library/items/22YIZBA8)) ([pdf](zotero://open-pdf/library/items/2UYJXF8N?page=7&annotation=YQRI3EPP)) -->

First I need to show that technological change has a direction.
This explains that fundamentally innovation does not simply bring progress or advance society, but that it takes us in different directions, or different configurations of sociotechnical systems [@andersson2021OutcomesDirectionalityMorphology;@yang2022ShapingDirectionalitySustainability].

Directed technological change @aghion2016CarbonTaxesPath, @moscona2022DoesDirectedInnovation. 
It is path-dependent, meaning events at one point in time effect later events, which may even lead to suboptimal outcomes [@david1985ClioEconomicsQWERTY].
Path-dependency arises due to positive feedback effects, such as increasing returns [@arthur1989CompetingTechnologiesIncreasing], or through extension of possibilities adjacent to exiting technology [@kauffman2000Investigations].
Instead of more or less progress, each innovation implies a new direction for sociotechnical systems [@stirling2011PluralisingProgressIntegrative].

@andersson2021OutcomesDirectionalityMorphology term the fact that some directions for technological change are more plausible than others _positive_ directionality.

, lending innovation _normative_ directionality [@andersson2021OutcomesDirectionalityMorphology].
As innovation systems are embedded in a spatial and temporal context, not all potentially desired pathways are available or feasible. Instead, sociotechnical systems are constrained by path-dependency [@kemp2022DiversifyingDeepTransitions;@bergek2015TechnologicalInnovationSystems; @stirling2011PluralisingProgressIntegrative], sometimes to the point of being locked-in to specific system states [@unruh2000UnderstandingCarbonLockin].

### Directing Technological Change

Then I will explain that with the normative turn of innovation policy the idea that policymakers should provide direction rather than just stimulating more innovation has spread.

As innovation policy took a normative turn under which innovation is needed to solve grand societal challenges [@schot2018ThreeFramesInnovation], the question of how to steer innovation into desired directions became a central issue in innovation research [@kohler2019AgendaSustainabilityTransitions].
Innovation policy now ought to promote the desired vector of societal development, instead of focusing on providing the right amount of progress [@schot2018ThreeFramesInnovation].
Dealing with the question of how to provide and exercise normative transition directionality @weber2012LegitimizingResearchTechnology synthesize policy insights from innovation systems and multilevel perspective research into a failures' framework.
Therein, directionality failure is understood as the failure to provide direction to the process of search and unclear or missing setting of collective priorities.
They caution that it should not be confused with shortsighted or flawed anticipation of market developments, as this myopia lacks a view on the direction of change.
To overcome this failure, that is, to provide directionality, they argue, external requirements need to be internalized by the innovation system in an interpretive process of negotiation between different system actors.
Reaching this collective understanding of direction, @weber2012LegitimizingResearchTechnology [p. 1042] write, is in practice, "... achieved by establishing shared future visions".
In their failures' framework, @weber2012LegitimizingResearchTechnology identify that the lack of (sufficiently clear) direction can represent a failure of the innovation system, which they termed directionality failure
<!-- of or in? -->.
To address this failure they proposed the need to generate collective visions for development outcomes.
<!-- Does it also include the means of achieving said outcomes? -->

Next I should explain who should provide directionality / decide.
Key authors are @mazzucato2018MissionorientedInnovationPolicies, @schot2018ThreeFramesInnovation, @parks2022DirectionalityTransformativeInnovation and @salasgirones2020RolePolicyEntrepreneurs.
Two central issues of normative directionality, or direction setting, are how desires might result in observable outcomes of new technological directions and, second, how to arrive at normative directionality in the first place.
On the theoretical side, @TOP-DOWN argue for the setting of directionality through a top-down process, while @BOTTOM-UP argue it should emerge from the bottom up.
Moving beyond the question of how normative directionality should emerge, @salasgirones2020RolePolicyEntrepreneurs studied how policy entrepreneurs attempt to influence the policymaking process to actively set directions for change in Dutch transport towards automated driving.
<!-- Empirically, the analysis has consisted of a multi-level case study that included the Swedish Innovation Agency, a strategic innovation programme, and a project organized by regional and urban actors. The analysis has shown how directionality was avoided in favour of demand articulation, allowing urban infrastructure companies alone to ‘give directions’ and articulate societal challenges. By excluding users and civil society, this process left little room for alternative ideas or conflicts that could encourage higher-order learning or experiment with radical solutions to societal challenges. The article’s findings indicate that policy layering has both advantages and disadvantages in transformative innovation policy. A disadvantage was how it resulted in an innovation programme led by actors from the supply side of innovation processes. These actors are not necessarily well-suited to organize innovation processes that address societal challenges, where it is necessary to open-up innovation to include stakeholders from the consumption and end-use of socio-technical networks such as urban infrastructure. Neither do these actors have the capacity to provide directionality to innovation processes. Yet at the urban and regional level, policy layering had the advantage of engaging with actors responsible for regional economic development. Their capacity to initiate innovation projects facilitated the participation of actors from the demand side, who might otherwise lack the capacity and networks to initiative innovations projects on their own. In this sense, existing institutions in innovation systems can contribute to societal challenges by helping to open-up innovation processes to actors from the demand side. Perhaps, if provided with models of innovation other than procurement, these existing institutions might facilitate more radical innovation that could encourage more transformative change. @parks2022DirectionalityTransformativeInnovation [11] -->

Lastly I turn to the question of how normative directionality is supposed to provide direction.
For this part I will use @stirling2024ResponsibilityHiddenPolitics.
I somewhere need to explain that what Striling [-@stirling2008OpeningClosingPower;-@stirling2010MulticriteriaDiversityAnalysis; -@stirling2011PluralisingProgressIntegrative] means with directionality is different from what other authors understand [c.f. @andersson2021OutcomesDirectionalityMorphology].

#### Shared Visions are Central Instruments to Provide Direction to Sustainability Transitions
Dealing with the question of how to provide and exercise normative transition directionality @weber2012LegitimizingResearchTechnology synthesize policy insights from innovation systems and multilevel perspective research into a failures' framework.
Therein, directionality failure is understood as the failure to provide direction to the process of search and unclear or missing setting of collective priorities.
They caution that it should not be confused with shortsighted or flawed anticipation of market developments, as this myopia lacks a view on the direction of change.
To overcome this failure, that is, to provide directionality, they argue, external requirements need to be internalized by the innovation system in an interpretive process of negotiation between different system actors.
Reaching this collective understanding of direction, @weber2012LegitimizingResearchTechnology [p. 1042] write, is in practice, "... achieved by establishing shared future visions".

In the two central research streams of sustainability transitions, the innovation systems and the multi-level perspective literature, the importance of visions is highlighted.
@hekkert2007FunctionsInnovationSystems provide a framework of intervention in innovation on a systemic level, which highlights the importance of guidance of search as one of seven functions innovation system must provide.
They explain guidance of search as, "those activities within the innovation system that can positively affect the visibility and clarity of specific wants among technology users" [@hekkert2007FunctionsInnovationSystems, 423].
Changing preferences in society, decides which direction of technological development is most promising through market and government influence and, "… an interactive and cumulative process of exchanging ideas between technology producers, technology users, and many other actors" [@hekkert2007FunctionsInnovationSystems, 423]. 
@bauer2017TechnologicalInnovationSystems adapts the innovation system functions to the specific case of biorefineries in Sweden and define the influence on the direction of search as "the incentives and/or pressures for organizations to enter the technological field. These may come in the form of visions, expectations of growth potential, regulation, policy targets, standards, articulation of demand from leading customers, crises in current business, etc." (p. 4).
Using the multi-level perspective, @yang2022ShapingDirectionalitySustainability to highlight the importance of visions in directionality over transformations, stressing the importance of shared visions between niche and regime actors.
Similarly, @bolton2019GrandVisionsPragmatic illustrate how visions and pragmatic change combined in the development and evolution of a highly integrated European electricity system.

@rotmans2001MoreEvolutionRevolution even go so far to argue that transition management is based on appealing and imaginative long-term visions and @loorbach2010TransitionManagementSustainable argues that shared visions need to be developed and disseminated between transition stakeholders.
But, as @bolton2019GrandVisionsPragmatic show, visions alone are insufficient to provide directionality.
In addition to being appealing, visions must also be specific and actors need to buy into them [@grillitsch2019InnovationPolicySystemwidea].
In summary, visions can be seen as shared imaginaries, mental models and aspirations of desired transitions which are expressed through narratives in discourse between actors [@bauer2018NarrativesBiorefineryInnovation; @holmgren2022WhoseTransformationThis].
However, one transition can be characterized by multiple, competing visions, as is illustrated by the transition to a bioeconomy.

### Sweden's Forest Bioeconomy

At a high level, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources [@bugge2016WhatBioeconomyReview].
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders [@dieken2021MultitudesBioeconomiesSystematic]. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficient wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other ecosystem services such as forest biodiversity.
Resolving these tradeoffs is challenging, as goal conflicts between different development pathways make simultaneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

As one of the leading forestry nations in the world, Sweden perceives substantial economic and political potential in advancing the bioeconomy and hence aims to promote a transition towards it.
In a report commissioned by the Swedish Government, @formas2012SwedishResearchInnovation lays out the plan for a Swedish bioeconomy based on two central pillars.
It foresees an economy based on,

@@

While the official Swedish bioeconomy strategy is still under development, the content of the quotes above echos across policy documents and the national forest program [@fischer2020SwedishForestsBioeconomy].
These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread, dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic] and form the main industrial visions [@befort2020GoingDefinitionsUnderstand]. 
A third vision, centered around ecology and an economy compatible with the biosphere acts as a counterpoint.

@@

In summary, despite -- or perhaps because of -- bioeconomy lacking a strong definition, the term is used to refer to different concepts, with different aims and objectives, foci, normative assumptions and hence implications for society and environment [@bugge2016WhatBioeconomyReview]. The summarized visions express themselves as narratives, actor coalitions and discourse artifacts in attempts to provide normative directionality to the transition.
This normative directionality is expected to translate into positive directionality of technological change through innovation.

# Data and Methods {#sec-data-methods}
